Towards user-friendliness in proof assistants: automated strategies \textbf{via} algebraic effects and handlers

### 2020-09-18 paper/main.tex
addresses comments

\newcommand{\comment}[1]{\footnote{Comment:~#1}}
\usepackage{pdfpages}
\begin{document}

@@

The usability of proof assistants such as Coq, Agda, Isabelle/HOL and Twelf is
hard to measure: all of them require significant domain knowledge that
frequently matches a deeper understanding of Type Theory, the theory that
enables such assistants to exist. Coq and Isabelle/HOL are more commonly applied
\comment{I'm unsure what ``that frequently matches a deeper understanding of
  Type Theory'' means... I~assumed the domain knowledge you meant \emph{was}
  type theory, and perhaps the details of the specific proof assistant.}
outside their niche, and their programming style is sharply different from
mainstream programming and even functional programming. Granted, they come with
their own integrated development environment (IDE), which may make it easier for
students and newcomers. To our knowledge, no usability study has been
held for a proof assistant, however there exists studies that account for
informal ``usability'' criteria -- they appear to measure how fast users can
\comment{Usage of ``appear'' seems like you're gonna really shit on these studies?}
familiarise themselves with the system or IDE.

Folklore within the proof assistants community says that newcomers
find Coq hard to use due to a large library of tactics and difficulty reading
the code after its completion. Many Coq learning materials suggest that users
add comments for structural induction cases and inductive hypothesis. As
mentioned, no user studies were conducted to attest such hypotheses.
\comment{Double use of hypothesis with different denotations is confusing.}

Granted, proof assistants are a relatively new technology
and there are no guidelines or even an intuition on how such systems should
perform or what features should be available to the users. The closest model we
\comment{Not sure what you mean by ``perform'', do you mean ``work/function''?}
have are proofs by hand. However, following that model for proof assistants has
two main pitfalls: (a) proofs by hand are not widely taught outside of
university mathematics courses; and (b) proof by hand and automated proof
employ fundamentally different paradigms, where the latter requires
axioms and formulations to be encoded explicitly, commonly known as a ``pedantic
proof style''.

@@

The reasons we employ algebraic effects and handlers are numerous:
\begin{enumerate}
\item as proofs cannot be fully automated, all approaches that try to automate
  the process (e.g, proof search, SMT solver) may be non-deterministic or
  never find a solution. Therefore, the system should be able to handle
  ``impure'' computations and errors. Algebraic effects and handlers have
  well-defined semantics and provide a simple interface for performing effects.
  With them, we avoid indiscriminate effects that are often error-prone and
  intricate effects machinery such as monad transformers;
\item the semantics of effects and handlers accommodate composition of arbitrary effects and the
  composition of multiple handlers, which means users have the ability to weaken
  more general strategies into specific ones while maintaining the original untouched;
  \comment{Not sure what you mean by ``original'' here?}
\item effects and handlers have well-defined semantics, and it is the main feature of many new
  research languages, such as Koka, Eff, Frank and Effekt, which suggests a greater
  potential.
  \comment{You may need to explain that potential; does it result from the
    prediction that more programmers will be familiar with effects in the
    future, due to their popularity?}
\end{enumerate}

@@

\paragraph{Algebraic Effects and Handlers} Algebraic effects were introduced by
**citation, and their handlers were introduced later on by **citation. An
algebraic effect is defined by an effect signature that comprises a set of
operations, and handlers that denote said operations. A handler may interpret
operations at the compiler-level, which was formally named co-model, but
colloquially called ``top-level handler''. Handlers are the most general form to
denote a set of operation, and more specific classes of handlers also exist.
\comment{I'm pretty confused by the past few sentences.}

\paragraph{The Essence of Witch}
In Figure \ref{fig:prototype}, we sum up our \textit{witch} approach.
As for the syntax, we use \texttt{operation \{ params \} via handler}, which is an
improvement over \texttt{handler(operation, params)}, since effect handling is similar
to function application, but also carries effect information.
The user defines data types and functions as usual, and then uses a
\textit{witch} to prove properties about said definitions. The Examples
section in Figure \ref{fig:prototype} shows this in action. The examples are
simplified for the sake of clarity -- we avoid
more complex properties and complicated proofs. The proof of
commutativity of addition under natural numbers and of associativity of list
concatenation are shown, and use the three main effects: Solver, Proof Search
and Tactics. In the proof assistant literature, there exists no precise
definition of commonly used terms ``solver'', ``proof search'' and ``tactics''.
All these terms are used in different communities, and mean some variation of ``automatic strategy to
construct a term under certain constraints''. Here, however, we use the
\comment{Not technically ``construct a term'', e.g., F* and Liquid Haskell's use
  of SMT solvers never constructs any term.}
following definitions:
\begin{itemize}
  \item The Solver effect is used for integration with external solvers via IO;
    we believe should suffice for the user to write, e.g. \texttt{SMT}, and handlers should
    implement internal strategies to choose between the different theories
    supported by solvers. If the black-box approach to solvers presents itself a
    challenge, a specialisation of the handler is possible, e.g.
    \texttt{operation \{ params \} via Z3.QF-UFLIA}.
  \item The Proof Search effect is used for library-level algorithms; users may
    choose to implement their own algorithms using a limited set of
    meta-programming constructs that are handled at top-level\footnote{The
      meta-programming constructs are operations of the Typechecker effect
      whose handlers are not available for the user.}.
    \comment{You essentially mean ``reflection'' here, right? You may want to
      use that word, or ``quoting'', or whatnot, explicitly.}
  \item The Tactics effect is used for strategies that simplify the goal at
    least one step, and may not complete all proof goals. This style is mainly
    used in the proof assistant Coq. While we do not foresee using as many as many
    tactics implemented as in Coq, we believe tactics such as
    \texttt{omega}, \texttt{ring}, \texttt{Admitted} are useful to users of
    all levels of expertise.
    \comment{All three of these tactics are total, so you may want to use
      examples which can fail or return partial proofs, such as \texttt{eauto}.}
  \item Lastly, the Error effect is used for feedback to the user,
    since any of the strategies may fail. Error has two operations,
    \texttt{throw} and \texttt{trace}: the former notifies the user that a strategy
    has failed, while the latter registers\footnote{Internally, the
      Typechecker effect should have a tree that stores all currently
      saved traces.} completed sub-goals during the
    strategy's attempt to complete the proof.
\end{itemize}

\paragraph{Witch's Architecture}
A \textit{witch} sits between the typechecker and the interactive mode of a
proof assistant. It uses constructs provided by the typechecker and communicates
with the interactive mode about the state of the current attempted strategy.
\comment{Perhaps rephrase this sentence, as it's a bit unclear?}
The user may manipulate proof handlers by composing or writing their own.

\paragraph{Downsides of our Approach}
There are two main styles for writing proofs within a proof assistant: external
verification and internal verification. The latter encourages code
\comment{Consider using extrinsic versus intrinsic.}
that is \textit{correct-by-construction}, meaning only valid states can be
encoded and values carry all information necessary to make proofs trivial; while
the former encourages writing code as usual, and reconstructing information
needed for proofs only when necessary.
Witch's primarily focus is on external verification. We believe that newcomers
will struggle to write proof-aware code, since it departs strongly from
mainstream programming, even within the functional style. We expect users to
write their code as they normally would do and then prove necessary properties.
Witch does not support writing correct-by-construction code. As proofs
are erased during compilation-time in Juvix, Witch could be extended
to be indexed by proof objects and parameterised over any type, such as
\texttt{\{A: Type\} $\rightarrow$ x y: A $\rightarrow$ }
\includegraphics[height=0.02\textheight]{image/hat.eps} \texttt{ (x * y $\equiv$
  y * x) } where we make sure that \texttt{*} over \texttt{A} is commutative.
\comment{Sentence is confusing, please rephrase?}
However, it is unclear whether having user's code under a monad-like structure leads
to well-structured code.
\comment{Sentence is confusing, please rephrase?}

@@

\subsection{Improving Witch Further}
While it is hard to say whether we have achieved our goals before user
evaluations, Witch is a first step towards a future where proof assistants are
friendlier to users, particularly newcomers. The present work is in its
infancy, and there are many open problems regarding Witch's ability to guide newcomers
towards correct-by-construction software, which constitutes the main improvement
we intend for Witch.

We hypothesise possibility of using proof
reconstruction, therefore relocating Witch to an assistant that lives solely
within the interactive mode -- as opposed to a framework for proof
automation in user code.
\comment{You should rephrase this -- it isn't a matter of proof reconstruction
  per se, since most strategies generate proof objects in the compiler. Instead,
it's a matter of inserting these objects into the source files. Note that this
makes the proofs more brittle, since one advantage of tactics is that if your
data types change slightly, proofs don't necessarily break, but instead your
solver may compute a difference solution.}

>>>>

\usepackage{pdfpages}
\begin{document}

@@

The usability of proof assistants such as Coq, Agda, Isabelle/HOL and Twelf is
hard to measure: all of them require significant domain knowledge in type
theory, the theory that enables such assistants to exist. Coq and Isabelle/HOL
are more commonly applied outside their niche, and their programming style is
sharply different from
mainstream programming and even functional programming. Granted, they come with
their own integrated development environment (IDE), which may make it easier for
students and newcomers. To our knowledge, no usability study has been
held for a proof assistant, however there exists studies that account for
informal ``usability'' criteria -- they appear to measure how fast users can
familiarise themselves with the system or IDE.

Folklore within the proof assistants community says that newcomers
find Coq hard to use due to a large library of tactics and difficulty reading
the code after its completion. Many Coq learning materials suggest that users
add code comments for structural induction cases and inductive hypothesis. As
mentioned, no user studies were conducted to attest that such strategy improves
the perceived quality of Coq proofs.

Granted, proof assistants are a relatively new technology
and there are no guidelines or even an intuition on how such systems should
work or what features should be available to the users. The closest model we
have are proofs by hand. However, following that model for proof assistants has
two main pitfalls: (a) proofs by hand are not widely taught outside of
university mathematics courses; and (b) proof by hand and automated proof
employ fundamentally different paradigms, where the latter requires
axioms and formulations to be encoded explicitly, commonly known as a ``pedantic
proof style''.

@@

The reasons we employ algebraic effects and handlers are numerous:
\begin{enumerate}
\item as proofs cannot be fully automated, all approaches that try to automate
  the process (e.g, proof search, SMT solver) may be non-deterministic or
  never find a solution. Therefore, the system should be able to handle
  ``impure'' computations and errors. Algebraic effects and handlers have
  well-defined semantics and provide a simple interface for performing effects.
  With them, we avoid indiscriminate effects that are often error-prone and
  intricate effects machinery such as monad transformers;
\item the semantics of effects and handlers accommodate composition of arbitrary effects and the
  composition of multiple handlers, which means users have the ability to weaken
  more general strategies into specific ones while maintaining top-level
  handlers unmodified;
\item effects and handlers have well-defined semantics, and it is the main feature of many new
  research languages, such as Koka, Eff, Frank and Effekt, which suggests a greater
  potential due to their popularity.
\end{enumerate}

@@

\paragraph{Algebraic Effects and Handlers} Algebraic effects were introduced by
**citation, and their handlers were introduced later on by **citation. An
algebraic effect is defined by a set of
operations, and handlers that denote said operations. A handler may interpret
operations at the compiler-level, which was formally named co-model, but
colloquially called ``top-level handler''. Handlers are the most general form to
denote a set of operations, and less general classes of handlers also exist.

\paragraph{The Essence of Witch}
In Figure \ref{fig:prototype}, we sum up our \textit{witch} approach.
As for the syntax, we use \texttt{operation \{ params \} via handler}, which is an
improvement over \texttt{handler(operation, params)}, since effect handling is similar
to function application, but also carries effect information.
The user defines data types and functions as usual, and then uses a
\textit{witch} to prove properties about said definitions. The Examples
section in Figure \ref{fig:prototype} shows this in action. The examples are
simplified for the sake of clarity -- we avoid
more complex properties and complicated proofs. The proof of
commutativity of addition under natural numbers and of associativity of list
concatenation are shown, and use the three main effects: Solver, Proof Search
and Tactics. In the proof assistant literature, there exists no precise
definition of commonly used terms ``solver'', ``proof search'' and ``tactics''.
All these terms are used in different communities, and mean some variation of
``automatic strategy to construct a term under certain constraints''. Here,
however, we use the following definitions:
\begin{itemize}
  \item The Solver effect is used for integration with external solvers via IO;
    we believe should suffice for the user to write, e.g. \texttt{SMT}, and handlers should
    implement internal strategies to choose between the different theories
    supported by solvers. If the black-box approach to solvers presents itself a
    challenge, a specialisation of the handler is possible, e.g.
    \texttt{operation \{ params \} via Z3.QF-UFLIA}.
  \item The Proof Search effect is used for library-level algorithms; users may
    choose to implement their own algorithms using a limited set of
    meta-programming\footnote{By meta-programming, we mean ``code that manipulates
    itself'', and not ``programming that happens in a higher level of abstraction''.
    For dependently typed programming languages, the usual term is reflection.
    However, we prefer not use reflection since it has another meaning in
    terms of effectful computations.} constructs that are handled at top-level\footnote{The
      meta-programming constructs are operations of the Typechecker effect
      whose handlers are not available for the user.}.
  \item The Tactics effect is used for strategies that simplify the goal at
    least one step, and may not complete all proof goals. This style is mainly
    used in the proof assistant Coq. While we do not foresee using as many as many
    tactics implemented as in Coq, we believe tactics such as
    \texttt{eauto}, \texttt{??}, \texttt{??} are useful to users of
    all levels of expertise.
    \item Lastly, the Error effect is used for feedback to the user,
    since any of the strategies may fail. Error has two operations,
    \texttt{throw} and \texttt{trace}: the former notifies the user that a strategy
    has failed, while the latter registers\footnote{Internally, the
      Typechecker effect should have a tree that stores all currently
      saved traces.} completed sub-goals during the
    strategy's attempt to complete the proof.
\end{itemize}

\paragraph{Witch's Architecture}
A \textit{witch} sits between the typechecker and the interactive mode of a
proof assistant. It uses meta-programming constructs provided by the typechecker and notifies
the interactive mode about whether the current attempted strategy was successful.
The user may manipulate proof handlers by composing or writing their own.

\paragraph{Downsides of our Approach}
There are two main styles for writing proofs within a proof assistant: external
verification and internal verification. The latter encourages code
that is \textit{correct-by-construction}, meaning only valid states can be
encoded and values carry all information necessary to make proofs trivial; while
the former encourages writing code as usual, and reconstructing information
needed for proofs only when necessary.
Witch's primarily focus is on external verification. We believe that newcomers
will struggle to write proof-aware code, since it departs strongly from
mainstream programming, even within the functional style. We expect users to
write their code as they normally would do and then prove necessary properties.
Witch does not support writing correct-by-construction code. As proofs
are erased during compilation-time in Juvix, Witch could be extended
to be indexed by proof objects and parameterised over any type, such as
\texttt{\{A: Type\} $\rightarrow$ x y: A $\rightarrow$ }
\includegraphics[height=0.02\textheight]{image/hat.eps} \texttt{ (x * y $\equiv$
  y * x) }.

@@

\subsection{Improving Witch Further}
While it is hard to say whether we have achieved our goals before user
evaluations, Witch is a first step towards a future where proof assistants are
friendlier to users, particularly newcomers. The present work is in its
infancy, and there are many open problems regarding Witch's ability to guide newcomers
towards correct-by-construction software, which constitutes the main improvement
we intend for Witch. We hypothesise possibility of inserting into the source
code a  \textit{proof sketch} that contains sub-goals of interest for the user and guides the
typechecker to rebuild the derivation tree faster. This change would, therefore,
relocate Witch to an assistant that lives solely
within the interactive mode -- as opposed to a framework for proof
automation in user code. The implementation of such feature may present itself
as a challenge, however.
